---
title: "Project_Progess_Report"
author: "Richardson, Trevor K./ Liu, Chi-Yun"
date: "`r Sys.Date()`"
output: pdf_document
fontsize: 12pt
geometry: margin=1in
header-includes:
    - \usepackage{setspace}\doublespacing
---

```{r setup, include=FALSE}
# load library 
library(tidyverse)
library(ggplot2)
library("usmap")
library(tidycensus)
library(kableExtra)
```

# Hypothesis

* Hypothesis 1: Heart disease is higher in the "Southern States" of the United States
* Hypothesis 2: Heart disease is higher amongst African Americans versus all other race/ethnicity
* Hypothesis 3: Heart disease is higher amongst males versus female

  Our first hypothesis is that **heart disease is higher in the Southern States**, this is largely supported by data that show that rates of heart disease are typically higher in African Americans and a larger portion of African Americans live in the “Southern” states. Additionally, it is supported by the data that shows that rates of obesity are higher in these states. 

* The scholarly journal linked here: <https://www.ajconline.org/article/S0002-9149(05)01211-7/fulltext> shows that African Americans are a unique population in regards to heart disease and suffer disproportionately when compared to other populations. 

* The scholarly journal linked here: <https://www.sciencedirect.com/science/article/abs/pii/S1047279707001536> shows that African Americans in southern states suffer from stroke due to heart disease versus African Americans in other regions of the US. The article terms the southern states as the “stroke belt” indicating the stroke is more prevalent in the southern states.

In the following, we will start to tidy and explore our data to find some evidence.

# Load Data

  We are loading the dataset from <https://chronicdata.cdc.gov/Heart-Disease-Stroke-Prevention/Heart-Disease-Mortality-Data-Among-US-Adults-35-by/i2vk-mgdh>

# Data input
```{r dataIn, echo=TRUE}
# read in data
heart_disease <- read_csv(file = "Heart_Disease_Mortality.csv")

dim(heart_disease)
#59076 rows, 19 columns
```

# Explore the Data

  Here we will summarize and explore the data
```{r veiwData, echo=TRUE}
#first few row
head(heart_disease)
# structure of data
str(heart_disease)
#summary
summary(heart_disease)
```


```{r findNa, echo=TRUE}
heart_disease %>%
  summarize(across(everything(), ~sum(is.na(.)))) # check the NA's value, remove NA's value in further calculation.
```


# Tidy data

  When looking at the heart disease mortality dataset, it is a bit messy and we need to tidy the data to make it look cleaner.

  As for the rates of heart disease, there are two columns of note. “Data_value” and “Data_Value_Unit”. When using the unique function on Data_Value_Unit there is only one value “Per 100,000 population.” We can rename "Data_value" to "RatePer100Thous" (Rate per 100,000 population), and then remove (not select) "Data_Value_Unit" from our new dataset.
  

```{r tidyData, echo=TRUE}
heart_disease %>% 
  separate(`Location 1`, into = c("LocationLat", "LocationLon"), sep = ",") %>% 
  mutate(LocationLat = parse_number(LocationLat),
         LocationLon = parse_number(LocationLon)) %>% 
  select(-c(DataSource, Class, Topic,Data_Value_Unit, Data_Value_Type, Data_Value_Footnote_Symbol,Data_Value_Footnote,StratificationCategory1,StratificationCategory2,TopicID)) %>%  # select the variables that we may use
  rename(Gender = `Stratification1`, RatePer100Thous =`Data_Value`, RaceEthnicity = `Stratification2`, State = LocationAbbr , City = LocationDesc, fips_code =`LocationID`)-> #rename the column
  new_hrt
new_hrt
```


```{r dataTable, echo=FALSE}
# data kable for powerpoint
new_hrt %>% 
  rename(HrtDisRate = `RatePer100Thous`) -> new_hrt_table
knitr::kable(
  new_hrt_table[1:10, ],
  caption = "Data Table"
)->new_hrt_table1

add_footnote(new_hrt_table1, "Heart Disease Rate (HrtDisRate): rates per 100,000")
```


# Evidence for hypothesis

## Hypothesis 1
```{r hypo1, echo=TRUE}

### Sort the data by state

rate_by_state <- new_hrt %>%
  group_by(State) %>%
  summarise(StateRate = mean(RatePer100Thous, na.rm = TRUE))

## Graph the data by state to show state differences
Rate_by_state %>% 
  mutate(State = fct_reorder(State, StateRate)) %>%
  ggplot(aes(x=State, y = StateRate)) + geom_col(width = 0.9, position = "dodge") +
  labs(x = "State", y = "Rate Per 100,000", title = "Rate of Heart Disease by State") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))->us_state_rate
# ggsave("us_state_rate.png", plot = us_state_rate, height = 16.25, width = 26.32, units = "cm")

```

```{r}
## Since a graph is a bit difficult to interpret the data since there are 50 states graphing on a US map would be better to interpret the data
mapdata <- data.frame( rate = rate_by_state$StateRate, state =rate_by_state$State)
plot_usmap(data = mapdata, values = "rate", color = "yellow", labels = TRUE) + 
  scale_fill_continuous(name = "Rate per 100,000", label = scales::comma) + 
  theme(legend.position = "right")+
  labs(title = "Rate of Heart Disease by State")
```
## Data Driven Hypothesis

The top five highest heart diease mortality state is GU/MS/AL/OK/AR (four out of five in southern)

```{r fipCode}
data("fips_codes")
fips_codes %>% 
  unite(state_county, state, county) %>% 
  unite(fips, state_code, county_code, sep = "")->fips_codes
```



```{r filterMs, echo=FALSE}
new_hrt %>% 
  filter(State == "MS") %>% 
  unite(state_county, State, City)->ms_data
ms_data %>%  
  group_by(state_county) %>% 
  summarise(city_rate = mean(RatePer100Thous, na.rm = TRUE)) %>% 
  left_join(fips_codes, by = c("state_county" = "state_county")) %>% 
  as.data.frame() %>% 
  drop_na()-> ms_rate

ms_rate %>% 
  separate(state_county, into = c("state", "county"), sep = "_")-> ms_rate


# ms_rate %>% summarize(across(everything(), ~sum(is.na(.))))
# There is a NA value in the dataset. It is Mississippi (state). we don't need it here
# drop.na, to be a new dataframe---ms_rate 
# summary(ms_rate)
ms_rate_top_30 <- top_n(ms_rate, 30, city_rate)

# summary(ms_rate_top_30)
```


```{r msCounties, echo=FALSE}
ms_rate %>%
  mutate(county = fct_reorder(county, city_rate)) %>% 
  ggplot(mapping = aes(x = county, y = city_rate))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


```{r msTop30, echo=FALSE}
ms_rate_top_30  %>%
  mutate(county = fct_reorder(county, city_rate)) %>% 
  ggplot(mapping = aes(x = county, y = city_rate))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```



```{r msMap,fig.width = 4, fig.asp = .9}

ms_map <- plot_usmap(regions = "counties", include = "MS", data = ms_rate,
                     values = "city_rate", color = "blue") + 
  scale_fill_continuous(low = "blue", high = "yellow", 
                        name = "rates per 100,000", label = scales::comma)+
  labs(title = "Mississippi Heart Diease Mortality") + 
  theme(panel.background = element_rect(color = "black", fill = "white")) +
  theme(legend.position = "top")

ms_map$layers[[2]]$aes_params$size <- 3
print(ms_map)
# ggsave("MS_county.png", plot = ms_map, height = 16, width = 30, units = "cm")

```

```{r filterAl, echo=FALSE}
new_hrt %>% 
  filter(State == "AL") %>% 
  unite(state_county, State, City)->al_data
al_data %>%  
  group_by(state_county) %>% 
  summarise(city_rate = mean(RatePer100Thous, na.rm = TRUE)) %>% 
  left_join(fips_codes, by = c("state_county" = "state_county")) %>% 
  as.data.frame() %>% 
  drop_na()-> al_rate

al_rate %>% 
  separate(state_county, into = c("state", "county"), sep = "_")-> al_rate

# al_rate %>% summarize(across(everything(), ~sum(is.na(.))))
# There is a NA value in the dataset. It is Alabama (state). we don't need it here
# drop.na, to be a new dataframe---al_rate 
# summary(ms_rate)
```


```{r alMap, echo=FALSE}
al_map <- plot_usmap(regions = "counties", include = "AL", data = al_rate,
                     values = "city_rate", color = "blue") + 
  scale_fill_continuous(low = "blue", high = "yellow", 
                        name = "rates per 100,000", label = scales::comma)+
  labs(title = "Alabama Heart Diease Mortality") + 
  theme(panel.background = element_rect(color = "black", fill = "white")) +
  theme(legend.position = "top")

al_map$layers[[2]]$aes_params$size <- 3.5
print(al_map)
# ggsave("AL_county.png", plot = al_map, height = 16, width = 30, units = "cm")
```

```{r filterOk, echo=FALSE}
new_hrt %>% 
  filter(State == "OK") %>% 
  unite(state_county, State, City)->ok_data
ok_data %>%  
  group_by(state_county) %>% 
  summarise(city_rate = mean(RatePer100Thous, na.rm = TRUE)) %>% 
  left_join(fips_codes, by = c("state_county" = "state_county")) %>% 
  as.data.frame() %>% 
  drop_na()-> ok_rate

ok_rate %>% 
  separate(state_county, into = c("state", "county"), sep = "_")-> ok_rate

# ok_rate %>% summarize(across(everything(), ~sum(is.na(.))))
# There is a NA value in the dataset. It is Oklahoma(state). we don't need it here
# drop.na, to be a new dataframe---ok_rate 
# summary(ok_rate)
```

```{r okMap, echo=FALSE}
ok_map <- plot_usmap(regions = "counties", include = "OK", data = ok_rate,
                     values = "city_rate", color = "blue") + 
  scale_fill_continuous(low = "blue", high = "yellow", 
                        name = "rates per 100,000", label = scales::comma)+
  labs(title = "Oklahoma Heart Diease Mortality") + 
  theme(panel.background = element_rect(color = "black", fill = "white")) +
  theme(legend.position = "top")

ok_map$layers[[2]]$aes_params$size <- 3.5
print(ok_map)
# ggsave("OK_county.png", plot = ok_map, height = 16, width = 30, units = "cm")
```


## Hypothesis 2
```{r hypo2, echo=TRUE}
## the sample of each race in dataset seems on average? 9846*6=59076
Rate_by_ethnicity <- new_hrt %>% 
  group_by(RaceEthnicity) %>%
  summarise(EthnicityRate = mean(RatePer100Thous, na.rm = TRUE))

## Graph ggplot
ggplot(Rate_by_ethnicity, aes(x=EthnicityRate, y = RaceEthnicity)) + geom_col() +
  labs(y = "Race", x = "Rate Per 100,000", title = "Rate of Heart Disease by Race") 

```


```{r raceNgender, echo=FALSE}
new_hrt %>% 
  group_by(RaceEthnicity, Gender) %>%
  summarise(EthnicityRate = mean(RatePer100Thous, na.rm = TRUE))->Rate_ethnicity_gender

race_gender <- ggplot(Rate_ethnicity_gender, aes(x=EthnicityRate, y = RaceEthnicity, fill = Gender), position = "dodge") + geom_col() +
  labs(y = "Race", x = "Rate Per 100,000", title = "Rate of Heart Disease by Race & Gender")
# ggsave("race_gender.png", plot = race_gender, height = 16, width = 30, units = "cm")

```




```{r}
ms_data %>%
  ggplot(mapping = aes(x = RatePer100Thous , y = RaceEthnicity))+
  geom_col()
```



```{r}
al_data %>% 
  ggplot(mapping = aes(x = RatePer100Thous , y = RaceEthnicity))+
  geom_col()
```

```{r}
ok_data %>% 
  ggplot(mapping = aes( x = RatePer100Thous , y = RaceEthnicity))+
  geom_col()
```


## Hypothesis 3
```{r hypo3, echo=TRUE}

rate_by_gender <- new_hrt %>% 
  group_by(Gender) %>% 
  summarise(Rate = mean(RatePer100Thous, na.rm = TRUE)) 

ggplot(rate_by_gender, aes(x=Gender, y = Rate)) + geom_col() +
  labs(x = "Gender", y = "Rate Per 100,000", title = "Rate of Heart Disease by Gender") 
```


```{r overall, fig.width = 7.5, fig.asp = .62, echo=FALSE}
overall_us <- new_hrt %>% 
  filter(Gender == "Overall" & RaceEthnicity == "Overall")

ggplot(data = overall_us)+
  geom_boxplot(mapping = aes(x = RatePer100Thous))+
  xlab("Heart Disease Mortality Rate (per 100,000)")
```




























